Do larger catchments respond different to forest cover change? Re-analysing a global data set.

### 2021-11-14 paper/Forest_and_Water/Forest_and_Water.tex
updates on paper 14112021

  \begin{abstract}
  This is the abstract.

  It consists of two paragraphs.
  \end{abstract}

@@

In addition, additional variables added were the latitude and longitude
for the center of the watershed as an approximation of its spatial
location. Using this information annual average reference
evapotranspiration (E0) was extracted from \textbf{XXXXX} if a value of
E0 was not available from the original papers. For large watersheds,
this value, similar to annual average rainfall, is only an approximation
of the climate at the location.

The length of the study can be a variable influencing the change in flow
(e.g.~Jackson et al., 2005) and therefore, the length, starting data and
completion date of the different studies was extracted from the
references provided by Zhang et al. (2017).

@@

@@

Building on the analyses by Zhang et al. (2017) and Filoso et al.
(2017), and combining their conclusions, the main hypothesis to test is
that the change in streamflow is impacted by the change in forest cover.
However, this change is clearly modulated by the area under
consideration (affecting the length of the flowpaths Zhou et al.
(2015)), the length of the study (c.f. Jackson et al. (2005)) and
possibly the climate (as indicated by either Pa and E0 or latitude and
longitude Filoso et al. (2017); Zhou et al. (2015)).

However, there could be further confounding factors, which are eluded to
by Filoso et al. (2017):\\
- the type of analysis, i.e.~paired catchment studies, modelling, time
series analysis etc.\\
- the age of the study, assuming that historical studies might not have
had the ability to measure at the accuracy that currently is available
to researchers, or that more careful historical attention to detail in
field studies might have been lost more recently due to reductions in
research investment.

@@

The first model applied in this analysis is based on the main hypothesis
outlined above, can the change in streamflow be predicted from the
change in forest cover, modulated by area, the length of the study and
the climate.

\[\tag{1}
\begin{aligned}
\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + {climate~type} + \varepsilon
\end{aligned}\]

However, the overall skewed distribution of the predictant
(\(\Delta \%Q\)) is problematic, and this results in a skewed
distribution of the GAM model residuals, which violates the linear model
assumptions. As a result we transformed \(\Delta \%Q\) and
\(\Delta \%forest\) back to fractions (0 - 1) and log transformed using
\(log10(x + 1)\), where \(x\) is either \(\Delta Q\) or
\(\Delta forest\). This means that the model residuals approximate
\(\sim N(0,\sigma^2)\) and this results in the following equation:

\[\tag{2}
\begin{aligned}
log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + {climate~type} + \varepsilon
\end{aligned}\]

A second model included all the variables in the analysis from Zhang et
al. (2017) in one model:

\[\tag{3}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

In this model, no direct interactions are assumed, and the assumption is
that all continuous variables (such as Pa) can have a linear or
non-linear relationship with \(log10(\Delta Q)\). This means that a
smooth function \(s()\) is applied to the variable. To restrict the
smoothness of the fit, the smoothness factor \(k\) is restricted to a
value of 3 (Wood, 2006). This restriction was applied to smooth
variables throughout this paper and we have dropped this from the
notation in subsequent equations.

For the model in equation 3, we only used the data from Zhang et al.
(2017) to make sure that the additional watersheds added to the data set
did not influence the analysis. Given that in Zhang et al. (2017),
dryness (\(\frac{E0}{Pa}\)) is used to look at variations in the change
in flow, we also fitted the following model:

\[\tag{4}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

Subsequently, using the full data set, including the additional
watersheds and the additional variables the following two models were
fitted:

\[\tag{5}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
& s(begin_{year}) + s(length_{study}) + {forest~type} + \\ & {climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

\[\tag{6}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + \\ 
& s(Latitude) + s(Longitude) + s(begin_{year}) + s(length_{study}) + \\ 
& {forest~type} + {climate~type} + {assessment~type} + \\  
& {hydrologic~regime} + \varepsilon
\end{aligned}\]

The results were analysed to identify:\\
1. the significance of the different variables\\
2. the direction of the categorical or shape of the smooth variables

>>>>

  \begin{abstract}
  This is the abstract. It consists of two paragraphs.
  \end{abstract}

@@

In addition, additional variables added were the latitude and longitude
for the center of the watershed as an approximation of its spatial
location. Using this information reference evapotranspiration (E\_\{0\})
was extracted from the Global Aridity Index and Potential
Evapo-Transpiration (ET0) Climate Databasev2 (Trabucco and Zomer, 2018),
if a value of E0 was not available from the original papers. For large
watersheds, this value, similar to annual average rainfall, is only an
approximation of the climate at the location.

The length of the study can be a variable influencing the change in flow
(e.g.~Jackson et al., 2005), as for example, more mature plantations are
thought to have smaller impacts on flow. Therefore, the length of the
study calculate as the difference between the starting data and
completion date of the different studies was extracted from the
references provided by Zhang et al. (2017).

@@

Similar to Zhang et al. (2017), the ``dryness index'' was calculated as:
\[\tag{1}
\begin{aligned}
D = \frac{E0}{Pa}
\end{aligned}\]

@@

\begin{verbatim}
## 
## -- Column specification --------------------------------------------------------
## cols(
##   Watershed = col_double(),
##   Latitiude = col_double(),
##   Longitude = col_double(),
##   rvalue_1 = col_double()
## )
\end{verbatim}

\begin{verbatim}
## 
## -- Column specification --------------------------------------------------------
## cols(
##   site_name = col_double(),
##   lat = col_double(),
##   long = col_double(),
##   `Area(km2)` = col_double(),
##   `Forest typ` = col_character(),
##   C = col_character(),
##   rvalue_1 = col_double()
## )
\end{verbatim}

\begin{verbatim}
## Warning: Removed 7 rows containing missing values (position_stack).
\end{verbatim}

\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-3-1.pdf}

\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-4-1.pdf}

@@

The general model tested is

\[\tag{2}
\begin{aligned}
\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
\end{aligned}\]

In this model, no direct interactions are assumed. The changes in forest
cover contain both positive (forestation) and negative values
(deforestation). In Zhang et al. (2017), these changes were jointly
analysed, assuming the effect on the change in flow was linear and
non-hysteretic. However, the impact of an increase in forest cover can
be different from the same fractional decrease in forest cover.
Therefore all the change in forest cover data is converted to positive
values, and an additional column (\emph{sign\_\{forest cover\}}) is
added that indicates whether it was a forest cover increase or decrease.
A further assumption in the model is that all continuous variables
\(Z_i\) (such as annual precipitation (Pa)) can have a linear or
non-linear relationship with \(\Delta Q \%\). This means that a smooth
function \(s()\) is applied to the \(Z_i\) variables.

For the model in equation 2, we initially only used the data from Zhang
et al. (2017) to make sure that the additional watersheds added to the
data set did not influence the results. Subsequently the analysis was
repeated and the additionally identified watersheds were added.

More generally the results were analysed to identify:\\
1. the significance of the different variables\\
2. the direction of the categorical or shape of the smooth variables

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{description-of-the-data}{%
\subsection{description of the data}\label{description-of-the-data}}

The overall dataset contains 312 observations of changes in flow. The
overall distribution of changes in flow is highly skewed as is the
distribution of changes in forest cover and Area. The values of changes
in flow greater than 100\% and smaller than -100\% clearly create long
tails on the change in flow distribution. Note also the large number of
studies with 100\% forest cover reduction.

\begin{verbatim}
## Warning: Removed 7 rows containing non-finite values (stat_bin).
\end{verbatim}

\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-8-1.pdf}

\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-9-1.pdf}

This immediately shows that for the data related to forest decreases,
there is almost always a positive flow change. In other words, flow
almost always increased. However, for increases in forest cover, this is
not the case, and flow can both increase and decrease. However in both
cases the variability in the reported change in flow increases with the
increase in forest cover change.

\hypertarget{the-initial-relationship-between-change-in-forest-cover-and-streamflow}{%
\subsection{The initial relationship between change in forest cover and
streamflow}\label{the-initial-relationship-between-change-in-forest-cover-and-streamflow}}

Following Zhang et al. (2017), the first step is to use a linear
regression to investigate the percent change in flow as a result in the
percent change forestry and modulated by the direction of the change,
either an increase in forest cover, or decrease in forest cover.

\begin{verbatim}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign
## 
## Parametric coefficients:
##                     Estimate Std. Error t value Pr(>|t|)    
## (Intercept)           8.8466     6.5008   1.361    0.175    
## DeltaF_perc_pos       0.5364     0.1017   5.274 2.52e-07 ***
## Forest_Signincrease -32.4760     6.9367  -4.682 4.26e-06 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## 
## R-sq.(adj) =   0.16   Deviance explained = 16.6%
## GCV = 3278.3  Scale est. = 3246.8    n = 312
\end{verbatim}

\includegraphics{Forest_and_Water_files/figure-latex/model1-1.pdf}

\begin{verbatim}
## 
## Method: GCV   Optimizer: magic
## Model required no smoothing parameter selectionModel rank =  3 / 3
\end{verbatim}

\begin{longtable}[]{@{}ccc@{}}
\toprule
\begin{minipage}[b]{0.11\columnwidth}\centering
Length\strut
\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\centering
Class\strut
\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\centering
Mode\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.11\columnwidth}\centering
0\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
NULL\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
NULL\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

While the overall variance explained in this model is not high at 0.16,
it clearly supports the hypothesized relationship between the change in
forest cover and the change in flow. The model suggests that for every
1\% change in forest cover, on the average, the flow changes 0.5\%.
However the change in flow is different for forest cover decreases
compared to forest cover increases. In fact, forest cover increases
decrease flow by 32\% less than a similar decrease in forest cover
causes flow to increase. So roughly speaking, a 1\% forest cover
increase on the average decreases flow by \((1 - 0.32)*0.5\%\), while a
the percentage forest cover decrease will increase flow by 0.5\%.

It is however clear from the lack of explaining power, that there could
be confounding factors, as alluded to in the methods. The obvious ones
being catchment dryness and area (following Zhang et al. (2017)).

\begin{verbatim}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm
## 
## Parametric coefficients:
##                       Estimate Std. Error t value Pr(>|t|)    
## (Intercept)          2.323e+01  9.189e+00   2.528   0.0120 *  
## DeltaF_perc_pos      5.424e-01  1.030e-01   5.265 2.64e-07 ***
## Forest_Signincrease -3.324e+01  6.955e+00  -4.779 2.73e-06 ***
## Area_km2            -1.867e-06  1.322e-05  -0.141   0.8878    
## Pa_mm               -1.067e-02  4.944e-03  -2.158   0.0317 *  
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## 
## R-sq.(adj) =  0.168   Deviance explained = 17.9%
## GCV = 3269.3  Scale est. = 3217      n = 312
\end{verbatim}

\includegraphics{Forest_and_Water_files/figure-latex/model2-1.pdf}

\begin{verbatim}
## 
## Method: GCV   Optimizer: magic
## Model required no smoothing parameter selectionModel rank =  5 / 5
\end{verbatim}

\begin{longtable}[]{@{}ccccc@{}}
\toprule
\begin{minipage}[b]{0.30\columnwidth}\centering
~\strut
\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
Estimate\strut
\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
Std. Error\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
t value\strut
\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
Pr(\textgreater\textbar t\textbar)\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{(Intercept)}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
23.23\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
9.189\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
2.528\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.01197\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{DeltaF\_perc\_pos}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.5424\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.103\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
5.265\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
2.64e-07\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Forest\_Signincrease}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-33.24\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
6.955\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-4.779\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
2.733e-06\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Area\_km2}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-1.867e-06\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
1.322e-05\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.1412\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.8878\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Pa\_mm}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.01067\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.004944\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.158\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.03169\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

Including area and annual precipitation does not really improve the
overall explaining power of the model, in fact, annual precipitation
appears to be only a very small confounding factor, representing only a
-0.01/\% partial effect in the change in streamflow, holding all other
factors constant. In contrast to earlier reported studies (Filoso et
al., 2017; Zhang et al., 2017), watershed area has no effect on the
change in stream flow. This supports our approach (in contrast to Zhang
et al. (2017)) to consider watershed area as a continuous variable and
making no separation between larger and smaller watersheds The main
effects remain the change in forest cover and whether this is an
increase or decrease.

\hypertarget{the-effect-of-location-on-the-globe}{%
\subsection{The effect of location on the
globe}\label{the-effect-of-location-on-the-globe}}

\begin{figure}
\includegraphics[width=0.9\linewidth]{../../data/FAOET0data} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
\end{figure}

As indicated, a second hypothesis relates to whether there is a strong
spatial global gradient as captured by latitude and longitude. As the
global map (@ref(fig:global\_map)) shows, the distribution of case study
catchments covers multiple continents and shows some distinct clustering
in parts of the world. Of interest is whether the spatial clustering
also indicates a difference in response to forest cover change.

\begin{verbatim}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm + 
##     Latitude + Longitude
## 
## Parametric coefficients:
##                       Estimate Std. Error t value Pr(>|t|)    
## (Intercept)          3.504e+01  1.026e+01   3.415 0.000724 ***
## DeltaF_perc_pos      5.147e-01  1.027e-01   5.014 9.07e-07 ***
## Forest_Signincrease -3.793e+01  7.230e+00  -5.247 2.90e-07 ***
## Area_km2            -6.282e-06  1.327e-05  -0.474 0.636137    
## Pa_mm               -1.304e-02  5.055e-03  -2.580 0.010339 *  
## Latitude            -3.119e-01  1.255e-01  -2.484 0.013520 *  
## Longitude           -1.058e-02  3.965e-02  -0.267 0.789747    
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## 
## R-sq.(adj) =  0.183   Deviance explained = 19.9%
## GCV = 3230.5  Scale est. = 3158      n = 312
\end{verbatim}

\includegraphics{Forest_and_Water_files/figure-latex/model3-1.pdf}

\begin{verbatim}
## 
## Method: GCV   Optimizer: magic
## Model required no smoothing parameter selectionModel rank =  7 / 7
\end{verbatim}

\begin{longtable}[]{@{}ccccc@{}}
\toprule
\begin{minipage}[b]{0.30\columnwidth}\centering
~\strut
\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
Estimate\strut
\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
Std. Error\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
t value\strut
\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
Pr(\textgreater\textbar t\textbar)\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{(Intercept)}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
35.04\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
10.26\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
3.415\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.0007243\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{DeltaF\_perc\_pos}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.5147\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.1027\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
5.014\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
9.071e-07\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Forest\_Signincrease}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-37.93\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
7.23\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.247\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
2.898e-07\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Area\_km2}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-6.282e-06\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
1.327e-05\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.4736\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.6361\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Pa\_mm}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.01304\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.005055\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.58\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.01034\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Latitude}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.3119\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.1255\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.484\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.01352\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Longitude}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.01058\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.03965\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.2669\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.7897\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

This linear model shows that there is a significant gradient in the
Latitude and with annual average rainfall, with watersheds closer to the
equator having lower changes in the runoff compare to watersheds further
away from the equator. This suggests an influence of radiation, which
will be tested next. In addition, the model suggests an influence of the
annual average rainfall, with wetter watersheds having slightly lower
changes in runoff. The total explaining power of the model is still low
at 0.18 suggesting further confounding factors currently not included in
the model.

There is no relationship with Longitude, suggesting that the different
continents do not show a trend in the East-West direction.

\hypertarget{impact-of-the-dryness-index}{%
\subsection{Impact of the dryness
index}\label{impact-of-the-dryness-index}}

The previous analysis suggests an influence of radiation on
evapotranspiration, and most likely related to the dryness index, as
also indicated in Zhang et al. (2017). Increased evapotranspiration
could lead to drier catchments, unless balanced by rainfall (such as
possibly in the tropics). This model introduces the dryness index as a
linear variable and drops the annual average precipitation as a
variables.

\begin{verbatim}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Latitude + 
##     Longitude + Dryness
## 
## Parametric coefficients:
##                       Estimate Std. Error t value Pr(>|t|)    
## (Intercept)          6.031e-01  8.582e+00   0.070 0.944022    
## DeltaF_perc_pos      5.006e-01  1.036e-01   4.831 2.18e-06 ***
## Forest_Signincrease -3.898e+01  7.253e+00  -5.375 1.55e-07 ***
## Area_km2            -7.227e-06  1.317e-05  -0.549 0.583604    
## Latitude            -3.121e-01  1.249e-01  -2.500 0.012974 *  
## Longitude           -6.775e-03  3.958e-02  -0.171 0.864199    
## Dryness              1.370e+01  3.628e+00   3.776 0.000193 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## 
## R-sq.(adj) =  0.204   Deviance explained = 21.9%
## GCV = 3199.2  Scale est. = 3125.7    n = 305
\end{verbatim}

\includegraphics{Forest_and_Water_files/figure-latex/model4-1.pdf}

\begin{verbatim}
## 
## Method: GCV   Optimizer: magic
## Model required no smoothing parameter selectionModel rank =  7 / 7
\end{verbatim}

\begin{longtable}[]{@{}ccccc@{}}
\toprule
\begin{minipage}[b]{0.30\columnwidth}\centering
~\strut
\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
Estimate\strut
\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
Std. Error\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
t value\strut
\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
Pr(\textgreater\textbar t\textbar)\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{(Intercept)}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.6031\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
8.582\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
0.07027\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.944\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{DeltaF\_perc\_pos}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.5006\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.1036\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
4.831\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
2.178e-06\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Forest\_Signincrease}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-38.98\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
7.253\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.375\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
1.549e-07\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Area\_km2}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-7.227e-06\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
1.317e-05\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.5487\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.5836\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Latitude}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.3121\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.1249\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.5\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.01297\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Longitude}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.006775\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
0.03958\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.1712\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.8642\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.30\columnwidth}\centering
\textbf{Dryness}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
13.7\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
3.628\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
3.776\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
0.0001927\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

The results from this model confirm that dryness is a significant
confounding factor of the change in streamflow as function of the change
in forest cover change. In fact if the dryness index doubles
(remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2*Pa,
which is very dry), the change in runoff is \textasciitilde14\% greater.
However, more interesting, Latitude remains a significant predictor with
each degree in latitude causing an -0.31\% change in runoff. This
indicates that Dryness (i.e.~an increase in radiation) alone does not
explain the trend in the Latitude and some other unknown confounding
factor is captured by Latitude.

\begin{verbatim}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2) + 
##     s(Dryness) + length + Assessment_technique + Latitude
## 
## Parametric coefficients:
##                              Estimate Std. Error t value Pr(>|t|)    
## (Intercept)                  19.80962   20.90807   0.947  0.34433    
## DeltaF_perc_pos               0.34576    0.10727   3.223  0.00144 ** 
## Forest_Signincrease         -33.34803    7.26563  -4.590 7.06e-06 ***
## length                       -0.16503    0.26850  -0.615  0.53935    
## Assessment_techniqueEA, HM    6.97930   47.05195   0.148  0.88220    
## Assessment_techniqueHM        1.39799   16.63165   0.084  0.93308    
## Assessment_techniquePWE       4.12984   19.29598   0.214  0.83070    
## Assessment_techniquePWE, HM  -5.44499   49.38888  -0.110  0.91230    
## Assessment_techniqueQPW      -9.50504   26.72374  -0.356  0.72239    
## Assessment_techniqueSH        2.52746   18.97865   0.133  0.89416    
## Latitude                     -0.30515    0.09629  -3.169  0.00172 ** 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Approximate significance of smooth terms:
##             edf Ref.df     F p-value
## s(Area_km2)   1      1 0.513   0.474
## s(Dryness)    1      1 0.002   0.964
## 
## R-sq.(adj) =  0.185   Deviance explained = 22.2%
## GCV = 2181.8  Scale est. = 2072.8    n = 260
\end{verbatim}

\includegraphics{Forest_and_Water_files/figure-latex/model5-1.pdf}

\begin{verbatim}
## 
## Method: GCV   Optimizer: magic
## Smoothing parameter selection converged after 15 iterations.
## The RMS GCV score gradient at convergence was 0.0001863276 .
## The Hessian was positive definite.
## Model rank =  29 / 29 
## 
## Basis dimension (k) checking results. Low p-value (k-index<1) may
## indicate that k is too low, especially if edf is close to k'.
## 
##             k' edf k-index p-value  
## s(Area_km2)  9   1    1.06   0.825  
## s(Dryness)   9   1    0.86   0.015 *
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\includegraphics{Forest_and_Water_files/figure-latex/model5-2.pdf}
\includegraphics{Forest_and_Water_files/figure-latex/model5-3.pdf}
